discussion

### 2018-02-27 man/ms_0_intro.Rmd
update intro
update texts

# Intro

Flujo:

@@

- importancia de analizar las sequías ya que estas están aumentando

- Esto cobra mayor enfasis en zonas mediterráneas, donde además de la existencia de sequías, hay una gran historia de impacto antrópico (land-use change driver, el otro driver del cambio global). Rececientemente Doblas-Miranda et al. han realizado una revisión donde han visto la importancia de la sequías, no solo como perturbación aislada, sino también como sus interacciones, siendo el land-use change el otra gran factor (motor) importante a tener en cuenta. Por tanto,

En una revision [@DoblasMiranda2017] subraya la relevancia y el impacto de las interacciones entre los emerging global change factors in the Mediterranean area. De entre las interacciones revisadas, la sequía es un factor crucial ...

past land uses could be a key factor altering the effects of current global changes and thus differentiating the Basin from other Mediterranean regions of the world

Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011]. Drought is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].

El estudio de como las sequías afectan a la vegetación es importante, sobre todo teneindo en cuenta que existen algunos estudios que sugieren un incremento de la severidad, extensión geográfica y frecuencia de las sequía a consecuencia del cambio climático [@Dai2011; @Trenberth2014]. Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg. @Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al. https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI). Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).

Junto con las sequías el cambio en el uso del suelo es otro de los grandes motores del cambio global.. 
... mejorar--> y esto puede ser especialmente importante en zonas como la región mediterránea,con un alto uso antrópico, donde la sequía presenta no solo efectos aislados, sino también la interacción con otros factores [@DoblasMiranda2017]. (ver abstract trabajo)

In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

@@

- Todo ello cobra especial importancia cuando las poblaciones de estudio están en su rear edge ... Por otro lado, es importante analizar las sequías en las poblaciones localizadas en el rear edge para ver su respuesta. (Jump et al)

@@

>>>>

# Introduction

@@

Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].

In the last decades there is an crescent interest in assessing the drought impacts on ecosystem, especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. There are studies analyzing the response of vegetation to drought from global (*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016) to local scales (*e.g* @MartinezVilalta2016 for a revision). Some studies have focusing on the drought effects on Mediterranean species which representing their southernmost limit [@Pasho2011; @Camarero2011; @Herrero2014; @Linares2014; @SanchezSalguero2017; @Caminero2018].

Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a]. In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017]. The impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is specially relevant for areas with a long history of landscape modification such us Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. In a recent review, @DoblasMiranda2017 found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are especially relevant for populations located in the retreating or rear edge of their distribution where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

There are several works que han analizado la respuesta de la vegetación a la sequía en su rear edge, y en este contexto es interesante analizar no solo el efecto de la sequía en estas poblaciones sino también su resiliencia. -- Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011). Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.

Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling.. // dorado-liñan // sanchez salguero

trabajos rear edge: 
* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
buscar otros

Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].

Remote sensing

These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al. 2009 , Matías and Jump 2015 ).

@@

### ojo un parrafito de resiliencia

Shifts in forest structure and composition in response to drought could alter ecosystem functions profoundly, affecting hydrological processes (Allen, 2007; Adams et al., 2012), nutrient cycling (Hughes et al., 2006; Xiong et al., 2011) and land-surface properties (Royer et al., 2011; Anderegg et al., 2013).

In this context, resilience is an important concept that is gaining unprecedented attention in ecology (e.g. Reyer et al., 2015). Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011). Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.

@@

Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg. @Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al. https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI). Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).

Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al. 2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al. ...)

### 2018-02-27 man/ms_3_discussion.Rmd
update texts

### OTRA COSA 
#### Climate and tree relationship 
Frase de resultados que tenemos que poner en discussion: 
Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ... **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.  
#### Disturbance
Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).

>>>>

### OTRA COSA

In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )

Environmental and forest management policies should take into account all these characteristics
of Mediterranean forests and the social, environmental and climatic conditions that are projected
for the coming years and decades.

#### Climate and tree relationship 
Frase de resultados que tenemos que poner en discussion: 
Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ... **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.  
#### Disturbance
Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
